Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-12 manuscript/manuscriptRW.Rmd
re-tructure of discussion following the new flow of ideas

Most of the models overestimate ingrowth rates, which is only
compensated by some models with high simulated mortality rates in the
early stages of population dynamics. The effect of light availability on
ingrowth rates is captured better than the effect of temperature and
soil moisture, but patterns are not consistent across models. Simulated
species diversity of ingrowth matches the observed ranges quite well.
Models simulating higher species diversity at the stand level do not
feature higher ingrowth diversity. The ingrowth of *Abies alba* is
captured across temperature and soil moisture gradients, whereas
ingrowth of *Fagus sylvatica* is generally underestimated. *Picea abies*
ingrowth is generally overestimated, particularly at colder sites.
Increasing complexity in the tree establishment modules of the models is
not related to higher accuracy of simulated tree establishment.
Furthermore, individual model design is more important than scale
(stand, landscape, global) and approach (empirical, process-based) to
accurately capture tree establishment.

@@

More specifically, we aimed to answer the following questions:

1)  Are models of forest dynamics capturing accurately tree ingrowth
    levels, initial tree diversity and mortality in tree establishment?

2)  Do model traits explain differences in model performance?

3)  How well do models capture total ingrowth and individual species
    regeneration niches across environmental gradients of light
    availability, temperature, and soil moisture?

@@

First, some models overestimated ingrowth across the entire gradients
for at least one of the five main tree species (Figure \@ref(fig:diff),
e.g., ForCEEPS, ForCEEPS(f), LandClim or LPJ-GUESS), while other models
overestimated ingrowth for more than one species for a large part of the
climate gradients, such as xComp, PICUS, SIBYLA or LPJ-GUESS (cf. red
colors in Figure \@ref(fig:diff)). These trends were consistent for both
ingrowth thresholds, although the exact changes in the share of the
ingrowth basal area were different (Figure S7).

Second, there was some congruence in the simulation results by species
across the models. The share in the ingrowth of Silver fir (*Abies
alba*) was captured well across the gradient by most models except for
xComp, PICUS, LandClim and Landis-II, which overestimated the ingrowth
share of this species. There were also some peculiarities evident for
some models, such as 4C that did not simulate *Abies alba*. The patterns
across models for *Fagus sylvatica* ingrowth were more complex, the
ingrowth was underestimated by many models across the environmental
gradients featuring negligible ingrowth at most sites, while others
overestimate it in different parts of the environmental space. Most of
the models overestimated the share of *Picea abies* ingrowth in the
cold-wet part of the gradients. *Pinus sylvestris* and *Quercus* spp.
represented a smaller share of the ingrowth in the observed data mostly
at warm-dry sites. Many models, however, erroneously featured ingrowth
for these two species along most of the climatic gradients although at a
low basal area share. Five models (4C, FORMIND, SIBYLA, xComp, and
LandClim) had almost no ingrowth of *Pinus sylvestris* and *Quercus*
spp., or did not feature any ingrowth of these species at all, close to
the patterns from the observed data.

Lastly, there is no model that performed well across the five species
regarding the species-specific ingrowth levels in the environmental
space of Figure 6. Some models (e.g., ForClim11, PICUS) tended to
consistently overestimate the ingrowth share of all five species, but
most models overestimated the ingrowth share of some species while
underestimating the share of others (e.g., ForCEEPS, LPJ-GUESS).

@@

## Ingrowth levels

@@

## Model formulation did not affect model performance

ADD THESE POINTS HERE complexity, empirical v. PB, etc.

@@

## Species composition of ingrowth

@@

Yet, there were pronounced variations how the main tree species were
represented along the environmental gradients, in particular the
dominant species Abies alba, Fagus sylvatica and Picea abies. The
ingrowth rates were sampled from simulations in the equilibrium, and in
this state it is expected that non-climax species such as *Pinus
sylvestris* or *Quercus* would be of minor importance, or absent
[@klopcic2015]. Most models captured this low abundance, which is also
found in the empirical data. Thus, although the broad patterns are
matched by many models, improvements in the quantification of the
regeneration niche of the species are needed, but this cannot be done in
the absence of robust datasets.

@@

## Ingrowth mortality

@@

@@

Based on the results of this study, it is clear that models of forest
dynamics need a focus on their establishment modules. Increasing
complexity in establishment modules has been motivated by better process
understanding and enhancing of model accuracy [@bugmann2022;
@bugmann2019]. However, more complex models do not necessarily lead to
better projections and rather to a reduced transparency and predictive
power [@franklin2020]. The results from this study showed that
increasing complexity in the establishment modules did not represent an
improvement in the projections of ingrowth levels, species composition
or mortality at early tree stages. Thus, the question what level of
detail one has to use to model establishment processes remains open and
must be addressed in future research [@konig_tree_2022; @bugmann2022].

We recommend that the further development of the establishment modules
is implemented as additional features that can be traced back, as done
in this study for the model variants of ForClim and ForCEEPS, and that
model complexity and structure must always be connected with model
objectives [@albrich2020]. If it should be necessary to include more
detail in the establishment models, this will come with higher
parameterisation efforts, which will most likely lead to lower
generalization as the required data will have to be collected from
specific locations, as currently there is no general establishment
dataset available.

@@

# Data availability statement

The code and data that support the findings of this study are openly
available at zenodo and gitlab (links to be added)

@@

Funding of the workshop by COST Action CA19139 PROCLIAS (PROcess-based
models for CLimate Impact Attribution across Sectors), supported by COST
(European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Grant ETH-35 18-1).

>>>>

Most of the models overestimate ingrowth rates, which is only
compensated by some models with high simulated mortality rates in the
early stages of population dynamics. Simulated species diversity of
ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The effect of light availability on ingrowth rates
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. The ingrowth of *Abies alba*
is captured across temperature and soil moisture gradients, whereas
ingrowth of *Fagus sylvatica* is generally underestimated. *Picea abies*
ingrowth is generally overestimated, particularly at colder sites.
Increasing complexity in the tree establishment modules of the models is
not related to higher accuracy of simulated tree establishment.
Furthermore, individual model design is more important than scale
(stand, landscape, global) and approach (empirical, process-based) to
accurately capture tree establishment.

@@

More specifically, we aimed to answer the following questions: (1) Are
models of forest dynamics capturing accurately tree ingrowth levels,
initial tree diversity and mortality in tree establishment? (2)Do model
traits explain differences in model performance? (3) How well do models
capture total ingrowth and individual species regeneration niches across
environmental gradients of light availability, temperature, and soil
moisture?

@@

First, some models overestimated ingrowth across the entire gradients
for at least one of the five main tree species (Figure \@ref(fig:diff),
e.g., ForCEEPS, ForCEEPS(f), LandClim or LPJ-GUESS), while other models
overestimated ingrowth for more than one species for a large part of the
climate gradients, such as xComp, PICUS, SIBYLA or LPJ-GUESS (cf. red
colors in Figure \@ref(fig:diff)). These trends were consistent for both
ingrowth thresholds, although the exact changes in the share of the
ingrowth basal area were different (Figure S8).

Second, there was some congruence in the simulation results by species
across the models. The share in the ingrowth of Silver fir (*Abies
alba*) was captured well across the gradient by most models except for
xComp, PICUS, and LandClim, which overestimated the ingrowth share of
this species. There were also some peculiarities evident for some
models, such as 4C that did not simulate *Abies alba*. The patterns
across models for *Fagus sylvatica* ingrowth were more complex, the
ingrowth was underestimated by many models across the environmental
gradients featuring negligible ingrowth at most sites, while others
overestimate it in different parts of the environmental space. Most of
the models overestimated the share of *Picea abies* ingrowth in the
cold-wet part of the gradients. *Pinus sylvestris* and *Quercus* spp.
represented a smaller share of the ingrowth in the observed data mostly
at warm-dry sites. Many models, however, erroneously featured ingrowth
for these two species along most of the climatic gradients although at a
low basal area share. Five models (4C, FORMIND, SIBYLA, xComp, and
LandClim) had almost no ingrowth of *Pinus sylvestris* and *Quercus*
spp., or did not feature any ingrowth of these species at all, close to
the patterns from the observed data.

Lastly, there is no model that performed well across the five species
regarding the species-specific ingrowth levels in the environmental
space of Figure \@ref(fig:diff)). Some models (e.g., ForClim11, PICUS)
tended to consistently overestimate the ingrowth share of all five
species, but most models overestimated the ingrowth share of some
species while underestimating the share of others (e.g., ForCEEPS,
LPJ-GUESS).

@@

## Most of the models overestimated tree ingrowth rates

@@

@@

@@

@@

@@

## Model traits did not affect model performance

Increasing complexity in establishment modules has been motivated by
better process understanding and enhancing of model accuracy
[@bugmann2022; @bugmann2019]. However, more complex models do not
necessarily lead to better projections and rather to a reduced
transparency and predictive power [@franklin2020]. The results from this
study showed that increasing complexity in the establishment modules did
not represent an improvement in the projections of ingrowth levels,
species composition or mortality at early tree stages. Thus, the
question what level of detail one has to use to model establishment
processes remains open [@konig_tree_2022; @bugmann2022].

## Regeneration niche

There were pronounced variations how the main tree species were
represented along the environmental gradients, in particular the
dominant species Abies alba, Fagus sylvatica and Picea abies. The
ingrowth rates were sampled from simulations in the equilibrium, and in
this state it is expected that non-climax species such as *Pinus
sylvestris* or *Quercus* would be of minor importance, or absent
[@klopcic2015]. Most models captured this low abundance, which is also
found in the empirical data. Thus, although the broad patterns are
matched by many models, improvements in the quantification of the
regeneration niche of the species are needed, but this cannot be done in
the absence of robust datasets.

@@

Based on the results of this study, it is clear that models of forest
dynamics need a focus on their establishment modules. As shown in the
results of the study it is still uncertain what level of detail one has
to use to model establishment processes and must be addressed in future
research. We recommend that the further development of the establishment
modules is implemented as additional features that can be traced back,
as done in this study for the model variants of ForClim and ForCEEPS,
and that model complexity and structure must always be connected with
model objectives [@albrich2020]. If it should be necessary to include
more detail in the establishment models, this will come with higher
parameterisation efforts, which will most likely lead to lower
generalization as the required data will have to be collected from
specific locations, as currently there is no general establishment
dataset available.

@@

# Reproducibility and data availability statement

This study compiles with the basic rules of reproducibility and has make
available the protocol of the study, the data and all the software that
leads of the findings presented in this study. The protocol is available
here XXX. The code and data that support the findings of this study are
openly available at zenodo and gihub (links to be added)

@@

Funding of the workshop was provided by the COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Grant ETH-35 18-1). ODY was funded by SNF project XXXX
